---
title: "Heat waves"
subtitle: "Current and Future Risks to Human Health"
author: "Brooke Anderson"
date: "March 7, 2016"
font: footnotesize
output:
  beamer_presentation:
    theme: "Boadilla"
---

## Biography

Training: 

- B.S., Chemical Engineering, N.C. State
- B.A., French Language and Literature, N.C. State
- Ph.D., Environmental Engineering, Yale University
- Postdoc, Dept. of Biostatistics, Johns Hopkins School of Public Health

## Biography

Research interests: 

- Health impacts of climate-related disasters (heat waves, hurricanes)
- Effects of air pollution on human health in China
- R open source software for "open" science

Service: 

- Editorial Board, *Epidemiology*
- Board of Associate Editors, *Environmental Health Perspectives*

Teaching: 

- R Programming

# Current heat risks: Heat and respiratory hospitalizations

## Heat and respiratory hospitalizations

\begin{block}{Primary study question}
Are daily Medicare hospitalizations for respiratory diseases
associated with increased summertime temperature?
\end{block}

\begin{block}{Additional study questions}
\begin{itemize}
\item Do effects differ by gender, age, or specific hospitalization
cause? 
\item Do observed effects change when the model includes air pollution?
\item What are the costs associated with heat-related hospitalizations?
\end{itemize}
\end{block}

## Study data

\begin{center}
213 US counties included in study.\\
\fbox{\includegraphics[width = 0.9\textwidth]{"figures/TemperatureMap"}} \\
\small{ Color of circle shows the county's
  daily mean temperature, averaged May--September, 1999--2008.}
  \end{center}
  
## Study model

\begin{block}{Statistical model}
  \vspace{-12pt}
  \begin{align*}
  log(\alert<1>{\mu_{t}^{c}}) \onslide<1->{=}
  &\onslide<2->{\alpha_{0}^{c} +}
  \onslide<2->{\alert<2>{\beta^{c}T_{t}^{c}} + }
  \onslide<3->{\alert<3>{\delta^{c}D_{t}} + }
  \onslide<4->{\alert<4>{f(H_{t}^{c})} +}
  \onslide<5->{\alert<5>{f(L_{t}^{c})}} \\
  & \onslide<6->{+ \alert<6>{B_{t}^{c}}}
  \end{align*}
  \vspace{-8pt}
\end{block}
  
\begin{block}{}
  \begin{equation*}
  \alert<1>{\mbox{Daily } \mbox{hospitalizations }} 
  \onslide<1->{= \mbox{function } \mbox{of }}
  \begin{cases}
  \onslide<2->{\mbox{\alert<2>{temperature}}}\\
  \onslide<3->{\mbox{\alert<3>{day of week}}}\\
  \onslide<4->{\alert<4>{\mbox{dewpoint}}}\\
  \onslide<5->{\alert<5>{\mbox{long-term trends}}}\\
  \onslide<6->{\alert<6>{\mbox{offset}}}
  \end{cases}
  \end{equation*}
\end{block}

## Combining county-level estimates for national estimate
  
\begin{block}{Two-level normal independent sampling estimation}
  \begin{align*}
  \hat{\beta^{c}} | \beta^{c}, \hat{\nu^{c}} &\sim \mathcal{N}(\beta^c,
  \hat{\nu^c}), c = 1, ..., 213 \\
  \beta^{c} | \phi, \tau^2 &\sim \mathcal{N}(\phi, \tau^2)
  \end{align*}
  where:
  
  \begin{description}
  \item[$\hat{\beta^c}$] Estimated temperature effect for county $c$
  \item[$\beta^c$] True temperature effect for county $c$
  \item[$\hat{\nu^c}$] Estimated variance of temperature effect for
  county $c$
  \item[$\phi$] True national temperature effect
  \item[$\tau^2$] Between-county variance in true temperature effects
  \end{description}
\end{block}

##

\begin{center}
\vspace{-1ex}
Temperature effects on Medicare respiratory hospitalizations.\\[1ex]
\fbox{\includegraphics[height = 0.87\textheight]{"figures/MainEffects"}}
\end{center}

## Effect modification by county climate
\begin{center}
County-level heat effects by average county temperature, May--September.\\[2ex]
\includegraphics[width = 0.75\textwidth]{"figures/TempVsEffect"}
\end{center}


## Study conclusions

\begin{itemize}
  \item Same-day respiratory hospitalizations increase significantly
  with hotter temperatures 
  \item These effects are similar for different respiratory causes, age
  groups, and genders 
  \item Effects are lower for cities with cooler climates 
  \item Effect estimates were stable to including ozone and particulate matter in the model
  \item Based on results, each $10^oF$ increase in temperature is associated with ~30 excess respiratory Medicare hospitalizations across the 213 study communities
\end{itemize}

## Other research on present risks of heat

\begin{block}{First / senior author} 
\begin{itemize}
\item Heat and risk of cardiorespiratory mortality in 107 U.S. communities
\item Modification of heat wave risks on mortality by heat wave characteristic in 43 US communities
\item County-level preparedness and response to 2011's extreme heat in the US
\item Impact of the 2003 power outage on human mortality in New York, NY
\end{itemize}
\end{block}

\begin{block}{Contributing author}
\begin{itemize}
\item Occupational heat-related mortality in the United States
\item Heat waves and mortality risk in Korea
\end{itemize}
\end{block}

# `futureheatwaves` package: Identifying, Characterizing, and Exploring Heat Waves in Climate Projections

## Patterns in heat waves with climate change 

\begin{center}
Patterns in heat wave length, frequency, and total days, 2061--2080\\
\fbox{
\includegraphics[width = 0.95\textwidth]{"figures/HeatwavePatterns"}
}
\end{center}

Source: [Oleson et al. 2015](http://www.theurbanclimatologist.com/uploads/4/4/2/5/44250401/olesonetal2015urbanruralheatcoldwaves.pdf)

## Multiple climate models

\begin{center}
Uncertainty across climate models
\fbox{\includegraphics[width = \textwidth]{"figures/WGI_AR5_Fig11-25"}}
\end{center}

Source: [IPCC Fifth Assessment, Working Group I, Figure 11-25](http://www.climatechange2013.org)

## Large-ensemble climate models

\begin{center}
Uncertainty across ensemble members from internal climate variability
\fbox{\includegraphics[width = \textwidth]{"figures/ClimateEnsemble"}}
\end{center}

Source: [Ball and Plant, 2008](http://rsta.royalsocietypublishing.org/content/366/1875/2603.short)

## `futureheatwaves` package

Design goals: 

- Ease processing of large sets of climate projections often needed for research on health impacts of heat waves under climate change
- Not only identify, but also characterize, all heat waves, to allow use of more complex epidemiological models
- Increase processing speed with C++
- Allow easy exploration across all resulting heat wave datasets

## `futureheatwaves` package

\begin{center}
Overview of `futureheatwaves` package.
\fbox{\includegraphics[width = \textwidth]{"figures/OverviewFigure"}}
\end{center}

## Heat wave definitions

\begin{center}
Examples of heat wave definitions used for health or climate studies.
\fbox{\includegraphics[width = \textwidth]{"figures/HeatwaveDefinitions"}}
\end{center}

Source: [Kent et al., 2014](http://ehp.niehs.nih.gov/1307262/)

## Heat wave definitions

```{r, eval = FALSE}
gen_hw_set(out = "example_results",
           dataFolder = projection_dir_location ,
           dataDirectories = list("hist" = c(1990, 1999),
                                  "rcp85" = c(2060, 2079)),
           citycsv = city_file_location,
           coordinateFilenames = "latlong.csv",
           tasFilenames = "projections.csv",
           timeFilenames = "timepoints.csv",
           IDheatwavesFunction = "new_function")
```

## Adaptation assumptions

\begin{center}
Relationship between "reference temperatures" and adaptation scenarios.
\fbox{\includegraphics[width = 0.9\textwidth]{"figures/AdaptationApproach2"}}
\end{center}

## Adaptation assumptions

```{r, eval = FALSE}
gen_hw_set(out = "example_results",
           dataFolder = projection_dir_location ,
           dataDirectories = list("hist" = c(1990, 1999),
                                  "rcp85" = c(2060, 2079)),
           citycsv = city_file_location,
           coordinateFilenames = "latlong.csv",
           tasFilenames = "projections.csv",
           timeFilenames = "timepoints.csv",
           referenceBoundaries = c(1990, 1999))
```

## Work on other packages

- `weathermetrics`: Convert between common weather metrics (including calculating heat index based on the National Weather Service's algorithm)
- `hurricaneexposure`: Create time series of tropical storm exposure based on rain, wind, and distance criteria
- `stormwindmodel`: Model tropical storm winds based on historical hurricane tracking data
- `countyweather`: Coordinate with `rnoaa` to pull observed weather data for county-level health studies

# Predicting high-mortality heat waves

## High-mortality heat waves

\begin{center}
Mortality during the 2003 French heat wave
\fbox{
\includegraphics[width = 0.9\textwidth]{"figures/HemonFrenchHeatwave"}
}
\end{center}

Source: [Pirard et al., 2005](http://www.eurosurveillance.org/ViewArticle.aspx?ArticleId=554)

## High-mortality heat waves

\begin{center}
High-mortality heat wave in Russia, 2010
\fbox{
\includegraphics[width = 0.8\textwidth]{"figures/RussiaHeatwave"}
}
\end{center}

Source: [pbs.org, 2010](http://www.pbs.org/newshour/rundown/heat-wave-wildfires-raise-deaths-in-moscow/)

## Predictive model of high-mortality heat waves

Applications: 

- (Short-term) Will next week have a high-mortality heat wave?
- (Mid-term) Which communities need to prioritize heat response plans?
- (Long-term) What are the health impacts and costs of severe heat waves under climate change scenarios?

## High-mortality heat waves

\begin{center}
Mortality risks during all heat waves in 82 US communities, 1987--2005
\fbox{\includegraphics[width = \textwidth]{"figures/HeatwaveDistribution"}}
\end{center}

## Predicting high-mortality heat waves

Category | Variable
---------|----------
Absolute intensity | Average $T_{mean}$
. | Highest daily $T_{mean}$
. | Lowest daily $T_{mean}$
Relative intensity | Quantile of average $T_{mean}$
. | Quantile of highest $T_{mean}$
. | Quantile of lowest $T_{mean}$
Timing | Day of year heat wave started
. | Month heat wave started
. | Whether heat wave was first in year

## Predicting high-mortality heat waves

Category | Variable
---------|----------
Length | Number of days
. | Days with $T_{mean} > 80^oF$
. | Days with $T_{mean} > 85^oF$
. | Days with $T_{mean} > 90^oF$
. | Days with $T_{mean} > 95^oF$
. | Days with $T_{mean} > 99^{th}$ perc.
. | Days with $T_{mean} > 99.5^{th}$ perc.
Community | Population
. | Population density
. | Long-term average $T_{mean}$
. | Long-term warm-season $T_{mean}$

## Predicting high-mortality heat waves

\begin{block}{Model types}
\begin{itemize}
\item Classification tree
\item Conditional tree 
\item Bagging
\item Random forest
\item Boosting
\end{itemize}
\end{block}

\begin{block}{Accounting for class imbalance}
\begin{itemize}
\item No adjustment
\item Oversampling from rare class
\item Over / under sampling
\item Randomly Over-Sampling Examples (ROSE)
\end{itemize}
\end{block}

## Model performance 

Model performance based on Monte Carlo cross-validation (all models use ROSE to account for class imbalance). \bigskip

Model | Recall | Precision (IQR) 
------|--------|-----------------
Classification tree | 94.0% | 2.6% (0.7%) 
Conditional tree | 87.5% | 7.2% (4.5%) 
Bagging | 94.0% | 2.6% (0.6%)
Random forest | 94.0% | 4.1% (2.0%) 
Boosting | 94.0% | 2.3% (0.5%)

## Variable importance

Most important variables across models (scaled to sum to 100% across each model). \bigskip

Variable | Class. tree | Bagging | Boosting
---------|-------------|---------|--------- 
Quantile of highest daily $T_{mean}$ | x | 64.44 | 64.60
Month heat wave started | x | 34.51 | 35.26
Average community warm $T_{mean}$ | | 0.33 | 0.00
Average community $T_{mean}$ | | 0.29 | 0.01
Number of days with $T_{mean} > 95^oF$ | | 0.27 | 0.12

## Heat wave characteristics

\begin{center}
Absolute intensity for high-mortality and less-dangerous heat waves
\fbox{
\includegraphics[width = 0.8\textwidth]{"figures/HeatwaveAbsTempDist"}
}
\end{center}

## Heat wave characteristics

\begin{center}
Relative intensity for high-mortality and less-dangerous heat waves
\fbox{
\includegraphics[width = 0.8\textwidth]{"figures/HeatwaveRelTempDist"}
}
\end{center}

## Multi-level structure

\includegraphics[width = \textwidth]{"figures/MultilevelHeatwaves"}

## "Out of sample" heat waves

\includegraphics[width = \textwidth]{"figures/HistoricalHeatwaves"}

## Improving the model?

- "Whole-pipeline" analysis
- Adding other variables: 
    - Power outages
    - Atmospheric patterns
    - Concurrent exposures (air pollution, humidity)
    
## Climate projection applications

- Avoided high-mortality heat waves under RCP8.5 versus RCP4.5
    - Collaboration with NCAR group
    - Used NCAR's large-ensemble model
- Uncertainty in projections of frequency of and exposure to high-mortality heat waves
    - Uncertainty related to climate model (CMIP5 models)
    - Uncertainty related to assumptions about adaptation