---
title: "Reproducible Research with R"
subtitle: "In-course exercise"
author: "Brooke Anderson"
date: "1/27/2021"
output: beamer_presentation
---

# Overview

## Overview

The in-class exercise will have two components: 

1. Some basic testing and exploration of creating and editing documents with
RMarkdown (10--15 minutes)
2. Start building a protocol for pre-processing data from plating to measure
the concentration of viable bacteria in samples (~ 25 minutes)

The second you will continue to work on with your group outside of class. You
will submit the result as your assignment for this week (due next week, I
believe).

## Groups

The groups for today are: 

- **Group 1:** Sere, Tyler, Tyler, Daniel, Pablo
- **Group 2:** Ikaia, Joey, Hector, Kaz, Naly
- **Group 3:** Camron, Emily, Sean, Carlos, Roland
- **Group 4:** Casey, Kristen, Alissa, Gaby, Lexi

# Basic exploration

## Working together in your group

- Plan to spend about 10--15 minutes on this part---it's really just to get a
feel for making and editing RMarkdown documents
- For this part, let those new to RMarkdown take the lead, and have one share
his or her screen.
- Team members who have used RMarkdown please guide those on your team who are
new and are trying this up while sharing their screen.

## Basic exploration of RMarkdown

For this first part of the exercise, do the following: 

1. Open RStudio. Create a new RMarkdown document that will render to 
Word.
2. Once the document opens, save the file, and then click on the 
"Knit" button to render the document to Word. You should get a new
output file in Word that's the result of rendering the RMarkdown.
3. Look at the RMarkdown document and the rendered Word document. See
if you can connect each part of the Word document to the part in the 
RMarkdown document that created it. 

## Basic exploration of RMarkdown

4. Change some things in the RMarkdown document. For example, delete 
some paragraphs or sections. Change some of the code in the code 
chunks. Use the RMarkdown cheatsheets---add some text with 
special formatting and try changing some code chunk options. Re-render
the document (with the "knit" button). 
5. Change something "by-hand" in the Word document. Re-render the 
RMarkdown document. Did it keep your change, or was it overwritten?
Discuss how this relates to the idea of "read only". 
6. Try using the green arrow in one of the code chunks to run the 
code just in that section. Try using the small arrow next to the 
"Knit" button to render the document to HTML, instead of Word. 

# Pre-processing protocol for plating data

## Pre-processing protocol for plating data

For the assignment, we are asking you to develop an RMarkdown document
that includes the code an description for how you **pre-processed**
plating data, to get from CFUs at different dilutions for a sample
to an estimate of the bacteria concentration in the sample.

- You will complete this assignment in the groups you are in today.
- Each person should submit their own copy of the group's final
document through Canvas (but we expect the final documents to 
be identical across group members). You should submit both the
RMarkdown file (".Rmd") and the output document (".docx", ".html", 
or ".pdf").

## Colony-forming units (CFUs)

You may be testing out some drugs against an infectious bacteria
and want to know how successful different drugs are in limiting
bacterial load. You run an experiment and have samples from 
animals treated with different drugs or under control.

You want to know: How much viable (i.e., replicating) bacteria are in 
each of your samples?

You can find out by **plating** the sample at different **dilutions** and
counting the **colony-forming units (CFUs)** that are cultured on each plate.

## Plating

You put a sample on a plate with a medium they can grow on and then give them
time to grow.

The idea is that individual bacteria from the original sample end up randomly
around the surface of the plate, and any that are viable (able to reproduce)
will form a new colony that, after a while, you'll be able to see.

## Dilution for counting

To count the number of colonies, you need **a "just right" dilution** (likely
won't know what this is until after plating) to have a **countable plate**.

If you have **too high** of a dilution (i.e., one with very few viable
bacteria), randomness will play a big role in the CFU count, and you'll
estimate the original with more variability. 

If you have **too low** of a dilution (i.e., one with lots of viable bacteria),
it will be difficult to identify separate colonies, and they may complete for
resources. (The pattern you see when the dilution is too low (i.e., too
concentrated with bacteria) is called a *lawn*---colonies merge together into
blobs.)

## Dilution for counting

To translate from diluted concentration to original concentration,
you need to do a back calculation. 

You have found: 

- The "just right" dilution
- The number of colonies that grew in that dilution

You need to use these numbers to estimate the concentration of viable 
bacteria in the original sample. 

## Example data

You will be working with example data from a package a
student at CSU (Amy Fox) is developing called 
`bactcountr` (https://github.com/aef1004/bactcountr). 

It provides data from plating samples from an experiment. The goal
is to estimate the concentration of *Mycobacterium tuberculosis* in 
two organs (spleen and lung) of mice.

## Example data

It is available to download at
https://github.com/aef1004/bactcountr/tree/master/data. Download this 
file and save it to the same directory as your RMarkdown for this 
assignment.

Because the data are in a binary R file, you can load it in your R session just
by using the `load` function (see next slide for example code).

## Example data

Here are the first few rows: 

```{r}
load("data/CFU_raw_formatted.rda")
head(CFU_raw_formatted, 3)
```

This dataset has already been converted to a "tidy" format, to make it easier to
work with in R. Each row gives the counted CFUs for one plate.

## Example data

It includes columns for: 

- `group`: The experimental group that the animal was assigned to
- `replicate`: The animal replicate. The rows with the same values for
*all* of group, replicate, and organ come from the same original sample.
- `organ`: Which organ of the animal the sample came from
- `dilution`: The level of dilution for the plated sample (with a 
dilution factor of 5, i.e., 5-fold dilutions at each step)
- `CFUs`: The colony-forming units (CFUs) counted on that plate.

Missing values in the `CFUs` for a certain dilution mean that the CFUs
were too numerous to count at that dilution.

## Example data

Since samples are plated at multiple dilutions, there are several rows per
original sample:

```{r message = FALSE, warning = FALSE}
library(tidyverse)
CFU_raw_formatted %>% 
  filter(group == 2 & 
           replicate == "A" & 
           organ == "Spleen")
```

## Pre-processing tasks

You'll need to do the following general pre-processing tasks with 
this data, describing and showing each in the RMarkdown
document for this week's assignment. 

1. Load the data into your R session
2. Explore the data, to show its dimensions and characteristics and
identify any potential quality control issues in the data
3. For each sample (measurement for one organ in one experimental animal), 
identify a good dilution to use for estimating the CFUs in the original 
sample
4. Use the CFUs counted for this "just right" dilution to estimate the
bacterial concentration in the original sample
5. An explanation of the final data's format (e.g., what columns
does it have, what's in each column)

## Pre-processing tasks

You may need to do some work to figure out a good way to do steps 3 and
4. 

Draw on any background knowledge you have. 

Also, feel free to use any
resource you can think of to figure out a reasonable way to do these
steps. 

Regardless of what you do and how you figure out how to do it, be sure
to document your choices clearly and thoroughly in the RMarkdown document. 

## Elements in the final document

Each document should include: 

- A meaningful title, the date the assignment is due, and the names of
all team members as authors (this should all be done in the YAML).
- Section headers for each section (there should be 5 sections, as 
listed above in the five pre-processing tasks)
- At least two references to journal articles or books to justify 
choices for how you've pre-processed the data. 
- At least one formatted mathematical equation / expression to explain 
your calculations
- Clear and thorough explanations for how you determined a good dilution
for each sample and how you calculated the concentration in the original 
sample using that dilution. 
- Minimal typos---use RMarkdown's spellcheck button!

## Pre-processing tasks

You may assume that: 

- The whole organ was used as the sample, and we are trying to 
estimate bacterial concentration in this whole organ
- In each case, the volume plated is 100 microliters
- The volume used to resuspend the CFU solution was 0.5 milliliters in 
each case
- The dilution factor is 5
- You may calculate either whole CFUs or log CFUs, just make sure you're
clear about which you present in your final data

If you need to make other assumptions, that's no problem---but make sure 
that you state any you make in the RMarkdown document so this is documented.


## Pre-processing tasks

Some `tidyverse` functions I found helpful: 

- `mutate`
- `abs` (takes an  absolute value) 
- `group_by` (you can group by more than one variable, 
so you can group all platings from the same original sample together)
- `arrange` (you can also arrange by more than one thing---it will only use
the later variables for arranging if there are ties based on earlier ones), 
- `slice`

## Resources

Resources on BibTex in RMarkdown: 

- https://bookdown.org/yihui/rmarkdown-cookbook/bibliography.html
- https://rmd4sci.njtierney.com/citing-articles-bibliography-styles.html
- https://rmarkdown.rstudio.com/authoring_bibliographies_and_citations.html

Resources on math in RMarkdown: 

- https://rmd4sci.njtierney.com/math.html
- https://rpruim.github.io/s341/S19/from-class/MathinRmd.html
- https://www.stat.cmu.edu/~cshalizi/rmarkdown/#math-in-r-markdown

