---
title: "Sample size determination"
author: "Brooke Anderson"
date: "9/26/2018"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, 
                      error = FALSE)
```

## Study goals

We are trying to determine if the sideshow incubators can double the
probability that a severely underweight infant survives. 

## Study design

We will study a cohort of severely underweight infants, some of whom will be
cared for at the sideshow in incubators (*treated group*) and some of whom
will be cared for without incubators (*control group*).

## Specify a hypothesis test on a parameter (along with the underlying probability model for the data)

For each infant in the study, we will know whether it is in the treated or
control group and whether it survives or dies. Let's define: 

- $\pi_1$: Proportion of infants who survive in the treatment group
- $\pi_2$: Proportion of infants who survive in the control group

In this case, unit of analysis for the study is an infant. The outcome
we will observe for each infant (survives [$Y = 1$] / dies [$Y = 0$]) is
expected to follow a *binomial distribution*. We now know enough to 
be pretty sure that our hypothesis testing will involve some kind of 
a test *comparing two proportions*.

Since we want to see if treatment doubles the probability of survival
(i.e., has a multiplicitive, rather than additive influence on the probability
of the outcome), we will conduct a hypothesis test on the *relative rate* 
(which we'll call $\lambda$),  which has the following relationship with these
proportions:

$$
\lambda = \pi_1/\pi_2
$$

We will conduct the following (two-sided, since we may also care about
evidence that the incubators decrease the probability of survival) hypothesis
test: 

$$
H_0 : \lambda = 1 
$$
$$
H_1 : \lambda \ne 1
$$

As a note, section 8.6 ("Testing a relative risk") in the chapter handout I
sent covers sample size calculations for this type of data and hypothesis
test. I am trying to use the same terminology as that section, to make
it easier to follow along with how this combines with the information in
the chapter. They set up the terminology for the hypothesis test above in 
a slightly different way, but you should be able to convince yourself 
that this set-up is identical to the hypothesis test they describe.

This chapter gives an equation to use for the sample size determination for
this type of hypothesis test and type of data (eq. 8.14). We will not
calculate this yet, but we will take a look so we'll know what other
parameters we'll ultimately need: 

$$
n = \frac{r + 1}{r(\lambda - 1)^2\pi^2} 
   \left[ z_\alpha \sqrt{(r+1) p_c (1 - p_c)} + 
   z_\beta \sqrt{\lambda\pi(1-\lambda\pi) + r\pi(1 - \lambda)} \right]^2
$$

where: 

- $r$ is the ratio of treated to control subjects in the study (i.e., $r = 1$
if you plan on having the same number of treated and control subjects)
- $\lambda$ is the effect size of scientific interest
- $\pi = \pi_2$ is the proportion of infants who survive in the control group
- $z_\alpha$ is a critical value that is set based on the significance level we select
- $p_c$ is the common proportion who survive over the two groups (see equation
below to estimate this)
- $z_\beta$ is a critical value that is set based on the target value of power for the test that we select

If you have equal sized groups (i.e., $r = 1$), you can estimate $p_c$ as: 

$$
p_c = \frac{\pi_1 + \pi_2}{2}
$$

We will use this information later to figure out all the parameters that
we'll need to set to determine the required sample size, and eventually we 
will use this equation to calculate the sample size.

## Specify the significance level $\alpha$ of the test

We decided on a significance level of $\alpha = 0.05$. 

Now that we've set that, we can determine that: 

$$
z_\alpha = 1.96
$$

Here's how you can deterine $z_\alpha$ from $\alpha$ in R (you're dividing
the $\alpha$ by 2 here because we're running a two-sided hypothesis test): 

```{r}
qnorm(1 - 0.05 / 2)
```

You could also use a z table, which you can find online or in many statistics
textbooks.

## Specify an effect size that reflects an alternative of scientific interest

Let's say that we've decided that the effect size we want to test is
a doubling in the probability of survival among severely underweight
infants. This translates to $\lambda = 2.0$. As we discussed in the 
group session, you may have good reasons to have selected a different
effect size of scientific interest here, which is fine as long as you 
justify that choice. 

## Obtain historical values or estimates of other needed parameters

Now, we're left with the following parameters that we still need to 
compute the power function of the test: 

- $r$ is the ratio of treated to control subjects in the study (i.e., $r = 1$
if you plan on having the same number of treated and control subjects)
- $\pi = \pi_2$ is the proportion of infants who survive in the control group
- $p_c$ is the common proportion who survive over the two groups (see equation below to estimate this)

First, let's plan to have the same number of infants in the treatment and 
control groups ($r = 1$). This choice of $r = 1$ will result in the most
power per study subject in most cases. When you decide to use a different
value for $r$, it may be related to practical constraints on your study. 
For example, you many only have a limited number of incubators, which would
limit the number of subjects you could treat, so in that case you may need
to plan to have several infants in the control group for every infant
in the treated group, which would result in a different value of $r$.

Next, we need to figure out a realistic value for $\pi_2$. In the group, 
we determined that a reasonable estimate for this is $\pi_2 = 0.2$ (i.e.,
20% of infants in the control group survive).

Finally, we need to calculate $p_c$. This is based on both $\pi_1$ and
$\pi_2$. Since we've specified that $\pi_1 = 0.2$ and that $\lambda = 2$, 
that requires $\pi_2 = 0.2 * 2 = 0.4$ (i.e., 40% of the treated infants 
survive). We can calculate $p_c$ as (since $r = 1$): 

$$
p_c = \frac{\pi_1 + \pi_2}{2} = \frac{0.2 + 0.4}{2} = 0.3
$$

## Specify a target value of the power of the test

Let's decide to set the power to 80%. Other values would be fine, too, 
as long as you have some justification for your choice. 

Based on this choice of power, $z_\beta$ = 0.8416.

In R: 

```{r}
qnorm(0.80)
```

## Calculate the sample size

We now have everything we need to calculate the required sample size. In R:

```{r}
# Define all the parameters for the equation
r <- 1
lambda <- 2.0
pi <- 0.2
z_alpha <- qnorm(1 - 0.05 / 2)
p_c <- (pi + pi * lambda) / 2
z_beta <- qnorm(0.80)

# Calculate the total required sample size (i.e., the sum of the
# required numbers in the treatment and control groups)
n <- ((r + 1) / (r * (lambda - 1) ^ 2 * pi ^ 2)) * 
  (z_alpha * sqrt((r + 1) * p_c * (1 - p_c)) + 
     z_beta * sqrt(lambda * pi * (1 - lambda * pi) + 
                     r * pi * (1 - pi))) ^ 2
n
```

Therefore, we should plan for a study with at least `r ceiling(n)` infants, 
half in the treatment group and half in the control group (you can 
round up if the total number is not even).

When we calculated this in OpenEpi, we got sample sizes of 164--182, 
depending on the method, so this result is consistent. 

## Setting up the analysis differently

The conclusions, in terms of required sample size, are approximately the same if
we define the outcome as death rather than survival. In this case, $p_2 = 0.8$
(i.e., an infant in the control group has an 80% probability of dying). If an
infant has twice the probability of surviving, it will have a 60% 
probability of dying, so in this case
$\lambda = 0.6 / 0.8 = 0.75$ (Tori, this is where we messed up in thinking
about this when we calculated this in OpenEpi).

```{r}
# Define all the parameters for the equation
r <- 1
lambda <- 0.75
pi <- 0.8
z_alpha <- qnorm(1 - 0.05 / 2)
p_c <- (pi + pi * lambda) / 2
z_beta <- qnorm(0.80)

# Calculate the total required sample size (i.e., the sum of the
# required numbers in the treatment and control groups)
n <- ((r + 1) / (r * (lambda - 1) ^ 2 * pi ^ 2)) * 
  (z_alpha * sqrt((r + 1) * p_c * (1 - p_c)) + 
     z_beta * sqrt(lambda * pi * (1 - lambda * pi) + 
                     r * pi * (1 - pi))) ^ 2
n
```

We come up with the same estimate of required sample size when we set the 
problem up this way.

## Another example

Here's an example of using R to calculate the results from Example 8.13 from 
the chapter: 

```{r}
# Define all the parameters for the equation
r <- 1
lambda <- 1.4
pi <- 0.02065
z_alpha <- qnorm(1 - 0.05) # One-sided test with alpha 0.05
p_c <- (pi + pi * lambda) / 2
z_beta <- qnorm(0.90)

# Calculate the total required sample size (i.e., the sum of the
# required numbers in the treatment and control groups)
n <- ((r + 1) / (r * (lambda - 1) ^ 2 * pi ^ 2)) * 
  (z_alpha * sqrt((r + 1) * p_c * (1 - p_c)) + 
     z_beta * sqrt(lambda * pi * (1 - lambda * pi) + 
                     r * pi * (1 - pi))) ^ 2
n
```

